---
title: "Processing land cover data from MAPBIOMAS Ecuador"
author: "Andres Marmol"
date: "2026-02-16"
output: html_document
---

```{r}
library(terra)
library(sf)
library(here)
library(dplyr)
library(stringr)
```

# 1. land cover recategorization

Given the more granular format of the Land cover file, it is necessary to bring it to the same scale as the environmental variables. As a first step, a 2 column matrix is created matching the original categories to those we want to use based on MapBiomas documentation.

```{r}

# matrix
rcl <- matrix(c(
  3,  3,
  4,  3,
  5,  3,
  6,  3,

  11, 12,
  12, 12,
  81, 12,
  82, 12,
  13, 12,
  29, 12,

  9,  21,
  21, 21,
  71, 21,

  24, 25,
  30, 25,
  23, 25,
  25, 25,
  68, 25,

  31, 33,
  33, 33,
  
  34, 34,
  
  27, 27
  
), ncol = 2, byrow = TRUE)

# defining codes and names
cat_codes <- c(3, 12, 21, 25, 33, 34, 27)
cat_names <- c("nat_forest", "non_nat_forest", "agriculture", "antropic", "water", "glacier", "not_observed")
```

# 2. Creating a function to upscale original land use files! 

## NOTE: this code will break if you do not update 'precip_ai' at the moment it is a raw layer, the environmental layers should be used instead.

```{r}
process_modal_file <- function(f, precip_ai, rcl, cat_codes, cat_names) {

  message("Reading: ", f)
  lu1 <- rast(f)

  # If values are stored as floats (e.g., 2.999999), force to integers
  lu1 <- round(lu1)

  # Limit to study area / template footprint (saves time & RAM)
  #lu1 <- crop(lu1, template30)
  #lu1 <- mask(lu1, template30)

  # Reclassify to your 5 categories (anything else -> NA)
  lu_cat <- classify(lu1, rcl = rcl, others = NA)

  # Proportion per 30" pixel for each category:
  # (0/1 mask at 1") -> resample average to 30" => proportion
  lu_cat30 <- rast(lapply(cat_codes, function(k) {
    m <- as.numeric(lu_cat == k)
    resample(m, precip_ai, method = "average")
  }))

  names(lu_cat30) <- cat_names

  # Keep NA footprint identical to template
  lu_cat30 <- mask(lu_cat30, precip_ai)

  # --------- Output naming ---------
  # Input pattern: (text1)_modal_(yy1)_(yy2).tif
  # Output name  : landcover_30arc_(yy1)_(yy2).tif
  base <- tools::file_path_sans_ext(basename(f))

  m <- regexec(".*_modal_(\\d{2})_(\\d{2}).*", base)
  yrs <- regmatches(base, m)[[1]]

  if (length(yrs) == 3) {
    yy1 <- yrs[2]
    yy2 <- yrs[3]
    out_file <- file.path(dirname(f), sprintf("landcover_30arc_%s_%s.tif", yy1, yy2))
  } else {
    # Fallback (won't crash if a filename doesn't match)
    out_file <- file.path(dirname(f), paste0(base, "_lu_cat30.tif"))
  }

  message("Writing: ", out_file)
  writeRaster(
    lu_cat30,
    filename = out_file,
    overwrite = TRUE,
    wopt = list(gdal = c("COMPRESS=DEFLATE", "PREDICTOR=2", "ZLEVEL=9"))
  )

  invisible(out_file)
}

```

# 3. Parsing all models with the name modal
Original files by Dome have been named modal. UPDATE: This has changed to list all the newly download files
```{r}
root_dir = "/Volumes/AM SSD/alianza_jambato/mapas mapbiomas"

modal_files <- list.files(
  root_dir,
  pattern="initiatives_ecuador_.*_\\d{4}\\.tif$",
  full.names = TRUE,
  recursive = TRUE,
  ignore.case = TRUE
)

message("found ", length(modal_files), " MapBiomas file(s).")
```

# 4. Running the loop

```{r}
out_files <- character(0)

for (f in modal_files) {
  out <- tryCatch(
    process_modal_file(f, precip_ai, rcl, cat_codes, cat_names),
    error = function(e) {
      message("FAILED: ", f, " | ", e$message)
      NA_character_
    }
  )
  out_files <- c(out_files, out)
}

out_files <- out_files[!is.na(out_files)]

message("Wrote ", length(out_files), " output file(s).")
out_files
```

# Additional information that might be needed but is not used in the current analyslis

These datasets have been already pre-processed. They all come from Chelsa Monthly Climate Data while Land Use comes from MapBIOMAS Ecuador 3.0. DEM has been extracted from LandSat (Confirm this with Dome).

A presing problem is that tiff files are not categorical, meaning that the interpretation of the results are not possible under this way. MapBiomas, unfortunately provides maps that are characterised as continuous categories, or at least that is how Terra and QGiS interprets them. Thus, they require a prior step before entering to the loop, as the numbers must be turned into categories, and named according its [documentation](https://ecuador.mapbiomas.org/wp-content/uploads/sites/7/2025/10/ATBD-Ecuador-Coleccion-3.0-ES.pdf#page=20.75 "MapBiomas Ecuador documentation").

here is the code:

```{r}
# calling raster
lu <- rast(here("Project", "Atelopus niche modelling", "single_SDM",
                "input_data", "aligned_modal_18_23_crop.tif"))

# extracting pixel values from tiff to object vals
vals = sort(unique(values(lu))) 

# entering category name for each of the values within the raster following MapBiomas Documentation
labs <- c("Forest", "Open forest", "Flooded forest", "Silviculture",
          "Non-Forest Wetland", "Grassland", "Other Non-Forest Natural formation",
          "Mosaics of Uses", "Antropic non-vegetated areas", "Not observed",
          "Rocky outcrop", "Mining", "Aquaculture", "River, lake and ocean", "Glacier")

# Transforming continuous values to categories (as.factor)
lu_cat <- as.factor(lu)

# creating dataframe with names for each value
lev <- data.frame(ID = vals, land_cover = labs)

# incorporating this to GEOtiff
levels(lu_cat) <- lev

# Writing the resulting tiff file
writeRaster(lu_cat, here("Project", "Atelopus niche modelling", "single_SDM", "input_data",
                         "aligned_LU_18_23_crop.tif"),
            overwrite = TRUE,
            wopt = list(gdal = c("COMPRESS=LZW")))

```