---
title: "Atelopus occurrences code"
author: "Andres Marmol & Amaru Rubio"
date: "2026-02-19"
output: html_document
---
This R Markdown notebook performs a Species Distribution Model (SDM) for *Atelopus ignescens* (the Jambato toad) using the biomod2 package. The workflow involves data preparation, pseudo-absence generation, Generalized Linear Model (GLM) fitting, and projecting the habitat suitability map.

This notebook works with relative pathways, meaning that the following structure for storing the data should be followed. For that the 'here' package is needed.



```{r}
## Load required libraries
require("terra")     # For handling spatial rasters and vectors
require("raster")    # (Legacy: Often still needed for compatibility with older spatial packages like biomod2)
require("sf")        # For simple features (vector spatial data)
require("biomod2")   # Core package for SDM
require("ggplot2")   # For enhanced plotting
require("dplyr")     # For data manipulation
require("tidyterra") # For plotting terra objects with ggplot2
require("ggtext")    # For advanced text formatting in ggplot2
require("here")      # Relative paths in this file
require("rgbif")     # GBIF R-pakage to pull occurence data from there.
require("lubridate") # package to edit data data formats.
require("randomForest")

## Set the working directory
# NOTE: Replace this path with your actual project directory.
```

# 1. Data Loading and Preparation

This section loads the species occurrence data directly from GBIF and the environmental predictor variables (rasters with the same extension, CRS and resolution).

### 1.1. Pulling GBIF occurrences

First, the taxonomic name of Atelopus must be confirmed.

```{r}
name_suggest(q="Atelopus ignescens", rank = "species")
```

Following, there is a check on the number of records available for the species, of which only those with coordinates are selected. At the end a dataframe called 'gbif_ai' is created. 'limit' is set up up to 2000 as GBIF contains 1637 records. Otherwise only 500 are retreived by default.

```{r}
gbif_ai <- occ_search(
              scientificName = "Atelopus ignescens",  # confirmed name of the species
              hasCoordinate=T,                        # logical argument to choose occ. with coordinates
              limit= 2000                             
              )

gbif_ai <- gbif_ai$data
```

### 1.2 Adding Alianza Jambato observations into the dataframe.

This dataset contains all the distribution points + collected over 2021-2022 by Alianza Jambato, Mateo Vega and Amanda Quezada.

```{r}
#project structure folder:
proj_root <- '/Alianza_Jambato_Git/Project/Atelopus niche modelling/single_SDM'

aj_ai <- read.csv(here("Project", "Atelopus niche modelling", "single_SDM",
                       "input_data", "occ_2018_2023.csv"))
aj_ai$countryCode = "EC"
aj_ai$stateProvince = "Cotopaxi"
```

```{r}
dfs = list(gbif = gbif_ai, df1 = aj_ai)

# creating a column indicating the df source
for (nm in names(dfs)) {
  dfs[[nm]]$source <- nm
}

# merging dfs by following columns in gbif_ai
all_cols = Reduce(union, lapply(dfs, names)) 

# a function to add missing colums and order them consistently
standardise_cols <- function(x, all_cols) {
  nr <- nrow(x)
  missing <- setdiff(all_cols, names(x))
  if(length(missing) > 0) {
    for (m in missing) x[[m]] <- rep(NA,nr) 
  }
  x[all_cols] # same order
}
df_ai <- lapply(dfs, standardise_cols, all_cols = all_cols)

# Combining dfs
dfaig <- do.call(rbind, df_ai)

# (optional) reset row names
row.names(dfaig) <- NULL
```

### 1.3. Checking duplicate coordinates and coordinates that are missplaced

Is always good practice to curate occurrence records from GBIF, as many of them could have innacurate identification of the corresponding species, or due to a lack of curation from the source (record hasn't been updated at the hosting institution, often a Natural History Museum). For that, we first use 'CoordinateCleaner' to clean GBIF data problems (e.g. point at province/states centroids) and finally a visual inspection based on literature review

```{r}
library("CoordinateCleaner")
library("countrycode")

# the package only works with ISO3c country codes (i.e ECU) and gfib is ISO2c (i.e. EC) thus it must be changed.

dfaig$countryCode <- countrycode(dfaig$countryCode,
                                 origin = "iso2c",
                                 destination = "iso3c")

dfaig_clean <- clean_coordinates(dfaig,
                                 lon = "decimalLongitude",
                                 lat = "decimalLatitude",
                                 countries = "countryCode",
                                 tests = c("centroids", "duplicates", "outliers"))
```

```{r}
plot(dfaig_clean)
```

```{r}
dfaig_cl <- dfaig[dfaig_clean$.summary,]
```

A few registers come from INaturalist and for protection reasons these are with a 30 km uncertainty. Thus, they need also to be filtered

```{r}
w <- "www.inaturalist.org"
m <- as.matrix(dfaig_cl)

rows <- apply(dfaig_cl, 1, function(r) any(grepl(w, r, ignore.case = TRUE)))
dfaig_cl <- dfaig_cl[!rows, ]

# there is an odd point to the east of napo tha twe are going to take out as well filtering for points -77.9

dfaig_cl <- dfaig_cl %>%
#dfaig_cl2 <- dfaig_cl %>%                  # interchange when you want to do the spThin
  filter(decimalLongitude <= -77.9)
```

### 1.4. Reduce biases in the dataset

#```{r}
library(spThin)

df_thinned <- spThin::thin(
  dfaig_cl2,
  lat.col = "decimalLatitude",
  long.col = "decimalLongitude",
  spec.col = "species",
  thin.par = 2, 
  reps = 50,
  locs.thinned.list.return = T, # creates tables based on repetitions.
  write.files = FALSE
) 

# extract the best thinned dataset
dfaig_cl <- df_thinned[[1]]
dfaig_cl$species = "Atelopusignescens"
#```


Lastly, occurences are comparted to the known literature by exploring them on a map

```{r}
library(geodata) 
```

```{r}
# GDAM Admin-1 (provinces) for Ecuador
ec <- geodata::gadm(country = "ECU", level = 1, path = "gadm_cache")

# Excluding Galapagos
ec_mainland <- ec[!grepl("Galápagos", ec$NAME_1, ignore.case = TRUE), ]

# choosing provinces with presence of A. ignescens based on BioWeb
prov <- c("Imbabura", "Chimborazo", "Cotopaxi", "Pichincha", "Tungurahua", "Napo", "Bolivar")
ai_prov <- ec[ec$NAME_1 %in% prov, ]
plot(ai_prov)
```

```{r}
plot(ec_mainland)
occ <- vect(dfaig_cl, geom = c("decimalLongitude", "decimalLatitude"), crs = "EPSG:4326")
#occ <- vect(dfaig_cl, geom = c("Longitude", "Latitude"), crs = "EPSG:4326") # activate only with spTHIN
occ <- project(occ, crs(ec_mainland))
plot(occ, add = TRUE)
```

filtering occurences based on provinces precense

```{r}
# CRS must match 
occ <- project(occ, crs(ai_prov))

# Keep only points inside the selected provinces
occ_sel <- terra::intersect(occ, ai_prov)
```

Cheking new distribution

```{r}
plot(ec_mainland)
plot(occ_sel, add = TRUE)
```

binging back the csv file from the occ spat vector created

```{r}

dfaig_cl <- as.data.frame(occ_sel, geom = "XY")
ai_occur <- data.frame(species = dfaig_cl$species, lon = as.numeric(dfaig_cl$x), lat = as.numeric(dfaig_cl$y))
ai_occur$occur <- 1 
```

```{r}
write.csv(dfaig_cl, file = here("Project", "Atelopus niche modelling", "single_SDM",
                       "input_data", "ai_occur_clean.csv"), col.names = TRUE)
```

While on travel
```{r}
dfaig_cl <- read.csv(here("Project", "Atelopus niche modelling", "single_SDM",
                       "input_data", "ai_occur_clean.csv"))
ai_occur <- data.frame(species = dfaig_cl$species, lon = as.numeric(dfaig_cl$x), lat = as.numeric(dfaig_cl$y))
ai_occur$occur <- 1 
```

