---
title: "climvar_ai_8590"
author: "Andres Marmol"
date: "2026-02-20"
output: html_document
---

# 1. set seeds for repetability

```{r}
#seedformat <- 10
#seedmodel <- 452
#seedproj <- 794
```

# 2. Defining the environmental predictors

### 2.1 Uploading DEM and cutting it
This layer was downloaded from earthexplorer.usgs.gov for a box containing Mainland Ecuador

Calling raw data
```{r}
DEM <- here("Project", "Atelopus niche modelling", "single_SDM", "input_data", "Ecuador_DEM", "10s090w_20101117_gmted_mea300.tif")
ecDEM <- rast(DEM)
```

cutting DEM to Atelopus ignescens presence based on provinces
```{r}
#checking coordinate systems
crs(ecDEM, describe = TRUE)
crs(ec_mainland, describe = TRUE)

# cropping and masking DEM
aiDEM_crop <- crop(ecDEM, ec_mainland)
aiDEM <- mask(aiDEM_crop, ec_mainland)

plot(aiDEM)

#writeRaster(aiDEM, filename = here("Project", "Atelopus niche modelling", "single_SDM", "input_data", "ai_DEM.tif"), overwrite = TRUE)
```
### 2.2 calling environmental and land use datasets

```{r}
# 2.1 Define Environmental Predictors
root_file <- here("Project", "Atelopus niche modelling", "single_SDM", "input_data")
input_files <- list.files(root_file, pattern = ".*2018-2021\\.tif$", full.names = TRUE) # change to 2018-2021 for the other date

# Read predictors (kept as you had it)
myExpl <- rast(input_files)

# Helper: crop + mask consistently
crop_mask <- function(x, v) mask(crop(x, v), v)

# Crop/mask predictors
myExpl <- crop_mask(myExpl, ec_mainland)

# Prepare DEM to match predictor grid
dem <- crop_mask(aiDEM, ec_mainland)

if (!same.crs(dem, myExpl)) dem <- project(dem, myExpl)
dem <- resample(dem, myExpl, method = "bilinear")

# Append DEM
myExpl <- c(myExpl, dem)

# Drop not_observed (handles "not_observed" and "not observed")
drop_names <- c("not_observed", "not observed", "glacier", "nat_forest", "non_nat_forest", "agriculture", "antropic", "water")
myExpl <- myExpl[[ !tolower(names(myExpl)) %in% drop_names ]]

# Final names
names(myExpl) <- c("pr", "tas", "tasmax", "tasmin", "DEM")

names(myExpl)
```

# 2.3 Variance Inflation factor (VIF)

```{r}
library(usdm)

# estimating Variance Inflated Factor (VIF) values
vif_values <- vif(myExpl)
print(vif_values)

# excluding values through stepwise process
vif_stepwise <- vifstep(myExpl, th = 10)
print(vif_stepwise)

# get the names of the safe vars
kept_vars <- vif_stepwise@results$Variables
print(kept_vars)

# names of the exluded variables
excluded <- vif_stepwise@excluded
print(excluded)
```
dropping variables based on the VIFs

```{r}
myExpl <- myExpl[[kept_vars]]

nlyr(myExpl)
names(myExpl)
```



# 3. Formatting the data to biomod2

first an overview of Atelopus occurence data points

```{r}
names(ai_occur)
str(ai_occur)
```

Now setting the data. Resp.var is the occurence data, often binary (0,1). resp.xy is linked to the coordinates of occurences and expl.var are all the climate, land use variables that will be assessed.

```{r}
aidata <- BIOMOD_FormatingData(
  resp.var = ai_occur[["occur"]],
  resp.xy  = ai_occur[, c("lon", "lat")],
  expl.var = myExpl,
  resp.name = "Atelopus.ignescens",
  PA.nb.rep = 2,
  PA.nb.absences = 500,
  PA.strategy = "random",
  filter.raster = FALSE,
  seed.val = seedformat
)
```
```{r}
plot(aidata)
```
# 4. Defining modelling options

I am running only a GLM, for its simplicity and interpretability. Other algorithms such as maxent, random forest, etc. can work but also have a higher chance of overfitting.

```{r}
aiopt <- bm_ModelingOptions(
  data.type = "binary",
  models = "GLM",
  strategy = "bigboss",
  bm.format = aidata
)
```
# 5. Run individual model

```{r}
aimodels <- BIOMOD_Modeling(
  bm.format = aidata,
  modeling.id = "Atelopus ignescens GLM",
  models = "GLM",
  OPT.user = aiopt,
  
  CV.strategy = "random",
  CV.nb.rep = 3,
  CV.perc = 0.7,
  
  metric.eval = c("AUCroc", "TSS"),
  
  seed.val = seedmodel,
  
  var.import = 3,
  nb.cpu = 2 
)
```

# 6. Model evaluation and optimal thresholds

```{r}
evals <- get_evaluations(aimodels)
evals
```


```{r}
# assesing the models through eval metrics

bm_PlotEvalMean(bm.out = aimodels, group.by = c("run"), dataset = c( "validation"))
bm_PlotEvalMean(bm.out = aimodels, group.by = c("run"), dataset = c( "calibration"))
```

finding the importance of the variables in the models
```{r}
ascores <- get_evaluations(aimodels)
vi <- as.data.frame(get_variables_importance(aimodels))

aggregate(var.imp ~ expl.var, vi, mean)
```

# building now the response plots
```{r}
ai_glm <- BIOMOD_LoadModels(aimodels, algo = "GLM")
```

```{r}
rc <- bm_PlotResponseCurves(
  bm.out         = aimodels,
  models.chosen  = ai_glm,
  new.env        = get_formal_data(aimodels, "expl.var"),
  #show.variables = get_formal_data(aimodels, "expl.var.names"),
  show.variables = c("pr", "tasmax"), # choose specific response variables
  fixed.var      = "mean",      # or "median"
  do.bivariate   = FALSE,
  do.plot        = TRUE
)
```
### 6.2 projecting the models

```{r}
models.proj <- get_built_models(aimodels, algo = "GLM")
# Project single models
aiproj <- BIOMOD_Projection(bm.mod = aimodels,
                                  proj.name = 'onlyclimate',
                                  new.env = myExpl,
                                  models.chosen = models.proj,
                                  metric.binary = 'all',
                            seed.val = seedproj
                                  )
```
# ###########################
```{r}
evals <- get_evaluations(aimodels)

tss_cutoff <- evals[evals$metric.eval == "TSS", "cutoff"]

summary(tss_cutoff)

# calculando desviación estandar y dispersion

sd_cutoff <- sd(tss_cutoff)
sd_cutoff
```


 Now we work with the model and projection rasters
```{r}
prob_raster <- get_predictions(aiproj)
model_names <- get_projected_models(aiproj)
evaltable <- get_evaluations (aimodels)

# extracting cutoffs 
current_cuttoffs <- evaltable[match(model_names, evaltable$full.name) &
                                evaltable$metric.eval == "TSS", "cutoff"]
current_cuttoffs <- as.numeric(current_cuttoffs)

# verificación de coincidencia con los rasters modelados
if(length(current_cuttoffs) != nlyr(prob_raster)) {
  stop("number of cuttoffs do not coincide with the number of models/rasters.")
}

# binary transformation
#aibinary <- bm_BinaryTransformation(data=prob_raster, threshold = current_cuttoffs)
aibinary <- bm_BinaryTransformation(data=prob_raster, threshold = 950)
plot(aibinary[[9]])
```


```{r}
thr <- c(950, 900, 800, 600, 500)

# Apply each threshold to the full 9-layer SpatRaster
bin_list <- lapply(thr, function(t) {
  b <- bm_BinaryTransformation(data = prob_raster, threshold = t)  # returns 9 layers
  b <- as.int(b > 0)  # ensure 0/1 (keeps NA as NA)
  names(b) <- paste0(names(prob_raster), "_thr", t)
  b
})

# Combine into one 36-layer SpatRaster
bin_stack <- rast(bin_list)

#Keeping only rastes for model with all runs and all data
keep <- c(9, 18, 27, 36, 45)
aibinary <- bin_stack[[keep]]

# Example: plot layer "originalName_thr950" (replace with a real layer name)
plot(aibinary)

names(aibinary)
```

```{r}
seedformat <- 10
seedmodel <- 452
seedproj <- 794
```


# Saving the rasters

```{r}
# cropping and masking to be only in reported provinces
aibinary_crop <- crop(aibinary, ai_prov)
aibinary_crop <- mask(aibinary_crop, ai_prov)

# turning 0 into no data:

# r is your multi-layer SpatRaster (stack)
aibinary_crop <- subst(aibinary_crop, from = 0, to = NA)

# writing rasters
writeRaster(aibinary_crop, 
      file = here("project", "Atelopus niche modelling", "single_SDM", "outputs", "aibin_vif_clim_1821.tif"),
      overwrite = TRUE,
      NAflag = -32768,
      datatype = "INT2S")

```


