---
title: "Raster processing for QGIS"
author: "Andres Marmol"
date: "2026-03-31"
output: html_document
---

This notebook reformats the binarised SDM outputs from both model variants —
climate-only (`aibin_vif_clim_1821.tif`) and all-variables
(`aibin_vif_allvar_1821.tif`) — into multi-band rasters suitable for direct
styled visualisation in QGIS.

Each source file contains five bands, one per suitability threshold (950, 900,
800, 600, 500 on the 0–1000 biomod2 scale, from most to least conservative).
The cascading `ifel()` logic assigns each pixel to the highest threshold band
at which it is classified as present, producing five non-overlapping presence
zones that can be stacked and styled in QGIS without overlap.

Run this notebook after both `clim_ai_1821.rmd` and `allvar_ai_1821.rmd` have
completed successfully.

# 1. Load libraries

```{r}
require("terra")  # Raster and vector spatial operations
require("here")   # Relative path management
```

# 2. Define a reusable processing function

The logic is identical for both model variants, so it is wrapped once in a
function that accepts an input file path, a label for plot titles, and an
output file path.

```{r}
process_for_qgis <- function(input_path, label, output_path) {

  if (!file.exists(input_path)) {
    stop(label, ": input file not found at ", input_path,
         "\nHas the corresponding SDM notebook been run?")
  }

  files <- rast(input_path)
  cat(label, "— layers found:", nlyr(files), "\n")
  cat("Layer names:", paste(names(files), collapse = ", "), "\n\n")

  # Inspect raw input
  plot(files, main = paste(label, "— raw input bands"))

  # Extract the five threshold bands (bands 1-5: 950, 900, 800, 600, 500)
  a <- files[[1]]  # threshold 950 — most conservative
  b <- files[[2]]  # threshold 900
  c <- files[[3]]  # threshold 800
  d <- files[[4]]  # threshold 600
  e <- files[[5]]  # threshold 500 — least conservative

  # Cascading ifel: each pixel assigned only to its highest present threshold
  band1 <- a
  band2 <- ifel(!is.na(b) &  is.na(a), 1, NA)
  band3 <- ifel(!is.na(c) &  is.na(b), 1, NA)
  band4 <- ifel(!is.na(d) &  is.na(c), 1, NA)
  band5 <- ifel(!is.na(e) &  is.na(d), 1, NA)

  out        <- c(band1, band2, band3, band4, band5)
  names(out) <- c("950", "900", "800", "600", "500")

  plot(out, main = paste(label, "—", names(out)))

  writeRaster(out,
              filename = output_path,
              overwrite = TRUE,
              NAflag    = -32768,
              datatype  = "INT2S")

  cat("Written to:", output_path, "\n\n")
  invisible(out)
}
```

# 3. Define input and output paths

```{r}
out_dir <- here("Project", "Atelopus niche modelling", "single_SDM", "outputs")

clim_in  <- file.path(out_dir, "aibin_vif_clim_1821.tif")
clim_out <- file.path(out_dir, "aibin_vif_clim_1821_qgis.tif")

allvar_in  <- file.path(out_dir, "aibin_vif_allvar_1821.tif")
allvar_out <- file.path(out_dir, "aibin_vif_allvar_1821_qgis.tif")
```

# 4. Process climate-only model

```{r}
process_for_qgis(
  input_path  = clim_in,
  label       = "Climate only",
  output_path = clim_out
)
```

# 5. Process all-variables model

```{r}
process_for_qgis(
  input_path  = allvar_in,
  label       = "All variables",
  output_path = allvar_out
)
```
