---
title: "Chelsa climate data process"
author: "Andres Marmol"
date: "2026-02-16"
output: html_document
---

# Intro

Climate datasets come in a resolution of 30 arc secs, in a CRS:WGS84 EPGS:4326. In a previous iteration, we tried to downscale the environmental layers to match the land cover layers, but this is an incorrect approach. While the grid continuously be subdivided, in reality we were only creating a pattern, which made no sense for the analysis.

This time we are keeping the original resolution of the environmental categories and upscale land cover for a better model performance.

```{r}
library(terra)
library(sf)
library(here)
```

# 1. Defining inputs

```{r}
# directories to work
dir_in <- '/Volumes/AM SSD/alianza_jambato/Chelsa_monthly'
dir_out <- here("Project", "Atelopus niche modelling", "single_SDM","input_data")
dir.create(dir_out, showWarnings = FALSE, recursive = TRUE)

#year ranges
yrs_6 <- 1985:1990
yrs_4 <- 2018:2021

vars <- c("pr", "tas", "tasmax", "tasmin")
#
```

# 2. defining the raster template

```{r}
# GDAM Admin-1 (provinces) for Ecuador
ec <- geodata::gadm(country = "ECU", level = 1, path = "gadm_cache")

# Excluding Galapagos
ec_mainland <- ec[!grepl("Galápagos", ec$NAME_1, ignore.case = TRUE), ]

# choosing provinces with presence of A. ignescens based on BioWeb
prov <- c("Imbabura", "Chimborazo", "Cotopaxi", "Pichincha", "Tungurahua", "Napo", "Bolivar")
ai_prov <- ec[ec$NAME_1 %in% prov, ]
plot(ai_prov)

# raster template:
precip <- rast('/Volumes/AM SSD/alianza_jambato/Chelsa_monthly/CHELSA_pr_03_1985_V.2.1.tif')
precip_ai <-crop(precip, ai_prov) # cuts to take into account the provinces where the jambato is
precip_ai <-mask(precip_ai, ai_prov) # masks the file
template <- precip_ai
```

# 3. listing files

```{r}
# listing files
files <- list.files(dir_in, pattern="\\.tif(f)?$", full.names=TRUE)
bn <- basename(files)

# identigying a name pattern text1_variable_month_year_text2-version.tif
rx <- "^([^_]+)_([^_]+)_(\\d{1,2})_(\\d{4})_(.+)\\.tif(f)?$"
m  <- regexec(rx, bn)
mm <- regmatches(bn, m)

ok <- lengths(mm) >= 6
files <- files[ok]
mm    <- mm[ok]

var   <- vapply(mm, `[[`, "", 3)
month <- as.integer(vapply(mm, `[[`, "", 4))
year  <- as.integer(vapply(mm, `[[`, "", 5))

# keep only target vars
keep <- var %in% vars
files <- files[keep]
var   <- var[keep]
year  <- year[keep]
month <- month[keep]

# sort 
ord <- order(var, year, month)
files <- files[ord]; var <- var[ord]; year <- year[ord]; month <- month[ord]
```

# 4. Creating the functions

## 4.1. Cropping and masking function

```{r}
# --- incremental mean with crop/mask first ---
mean_from_files_cropmask <- function(files, template, resample_method = "bilinear") {
  stopifnot(length(files) > 0)

  tmpl <- template
  s <- NULL
  n <- NULL

  for (f in files) {
    r <- rast(f)

    # 1) crop to bbox of template
    r <- crop(r, ext(tmpl))

    # 2) mask to template footprint
    r <- mask(r, tmpl)

    # 3) force exact grid match (only if needed)
    if (!compareGeom(tmpl, r, stopOnError = FALSE)) {
      r <- resample(r, tmpl, method = resample_method)
      r <- mask(r, tmpl)
    }

    # 4) running sum + count
    if (is.null(s)) {
      s <- r
      n <- !is.na(r)
      s[is.na(s)] <- 0
    } else {
      n <- n + (!is.na(r))
      r[is.na(r)] <- 0
      s <- s + r
    }
  }
  
  s / n
}
```

## 4.2. Mean function

```{r}
run_period <- function(v, years_vec, label) {
  f <- files[var == v & year %in% years_vec]

  if (length(f) == 0) {
    message("Skipping ", v, " for ", label, " (no files)")
    return(invisible(NULL))
  }

  rmean <- mean_from_files_cropmask(f, template, resample_method = "bilinear")

  out <- file.path(dir_out, sprintf("mean_%s_%s.tif", v, label))
  writeRaster(rmean, out, overwrite = TRUE)

  message("Wrote: ", out, " (n=", length(f), ")")
}
```

# 5. Processing all climate variables

```{r}
label_6 <- sprintf("%d-%d", min(yrs_6), max(yrs_6))
label_4 <- sprintf("%d-%d", min(yrs_4), max(yrs_4))

for (v in vars) {
  run_period(v, yrs_6, label_6)
  run_period(v, yrs_4, label_4)
}
```

# 6. Quality checks

```{r}
for (v in vars) {
  cat("\n", v, "\n")
  cat("1985-1990:", sum(var==v & year %in% yrs_6), "\n")  # expect 72 if complete
  cat("2018-2021:", sum(var==v & year %in% yrs_4), "\n")  # expect 48 if complete
}
```
```{r}
pr_85_90 <- rast(here("Project", "Atelopus niche modelling", "single_SDM","input_data", "mean_pr_1985-1990.tif"))
pr_18_21 <- rast(here("Project", "Atelopus niche modelling", "single_SDM","input_data", "mean_pr_2018-2021.tif"))
```



