---
title: "Processing land cover data from MAPBIOMAS Ecuador"
author: "Andres Marmol"
date: "2026-02-16"
output: html_document
---

```{r}
library(terra)
library(sf)
library(here)
```

# 1. land cover recategorization

Given the more granular format of the Land cover file, it is necessary to bring it to the same scale as the environmental variables. As a first step, a 2 column matrix is created matching the original categories to those we want to use based on MapBiomas documentation.

```{r}

# matrix
rcl <- matrix(c(
  3,  3,
  4,  3,
  6,  3,

  11, 12,
  12, 12,
  13, 12,
  29, 12,

  9,  21,
  21, 21,

  27, 25,
  30, 25,
  25, 25,

  31, 33,
  33, 33,
  34, 33
), ncol = 2, byrow = TRUE)

# defining codes and names
cat_codes <- c(3, 12, 21, 25, 33)
cat_names <- c("nat_forest", "non_nat_forest", "agriculture", "antropic", "water")
```

# 2. Creating a function to upscale original land use files! 

## NOTE: this code will break if you do not update 'precip_ai' at the moment it is a raw layer, the environmental layers should be used instead.

```{r}
process_modal_file <- function(f, precip_ai, rcl, cat_codes, cat_names) {

  message("Reading: ", f)
  lu1 <- rast(f)

  # If values are stored as floats (e.g., 2.999999), force to integers
  lu1 <- round(lu1)

  # Limit to study area / template footprint (saves time & RAM)
  #lu1 <- crop(lu1, template30)
  #lu1 <- mask(lu1, template30)

  # Reclassify to your 5 categories (anything else -> NA)
  lu_cat <- classify(lu1, rcl = rcl, others = NA)

  # Proportion per 30" pixel for each category:
  # (0/1 mask at 1") -> resample average to 30" => proportion
  lu_cat30 <- rast(lapply(cat_codes, function(k) {
    m <- as.numeric(lu_cat == k)
    resample(m, precip_ai, method = "average")
  }))

  names(lu_cat30) <- cat_names

  # Keep NA footprint identical to template
  lu_cat30 <- mask(lu_cat30, precip_ai)

  # --------- Output naming ---------
  # Input pattern: (text1)_modal_(yy1)_(yy2).tif
  # Output name  : landcover_30arc_(yy1)_(yy2).tif
  base <- tools::file_path_sans_ext(basename(f))

  m <- regexec(".*_modal_(\\d{2})_(\\d{2}).*", base)
  yrs <- regmatches(base, m)[[1]]

  if (length(yrs) == 3) {
    yy1 <- yrs[2]
    yy2 <- yrs[3]
    out_file <- file.path(dirname(f), sprintf("landcover_30arc_%s_%s.tif", yy1, yy2))
  } else {
    # Fallback (won't crash if a filename doesn't match)
    out_file <- file.path(dirname(f), paste0(base, "_lu_cat30.tif"))
  }

  message("Writing: ", out_file)
  writeRaster(
    lu_cat30,
    filename = out_file,
    overwrite = TRUE,
    wopt = list(gdal = c("COMPRESS=DEFLATE", "PREDICTOR=2", "ZLEVEL=9"))
  )

  invisible(out_file)
}

```

# 3. Parsing all models with the name modal
Original files by Dome have been named modal.
```{r}
root_dir = here()

modal_files <- list.files(
  root_dir,
  pattern="modal.*\\.tif(f)?$",
  full.names = TRUE,
  recursive = TRUE,
  ignore.case = TRUE
)

message("found ", length(modal_files), " modal file(s).")
```

# 4. Running the loop

```{r}
out_files <- character(0)

for (f in modal_files) {
  out <- tryCatch(
    process_modal_file(f, precip_ai, rcl, cat_codes, cat_names),
    error = function(e) {
      message("FAILED: ", f, " | ", e$message)
      NA_character_
    }
  )
  out_files <- c(out_files, out)
}

out_files <- out_files[!is.na(out_files)]

message("Wrote ", length(out_files), " output file(s).")
out_files
```

