---
title: "Species Distribution Model (SDM) for Atelopus ignescens"
Author: "Domenica G. Cevallos & Andrés Mármol-Guijarro
output:
  word_document: default
  html_notebook: default
---

This R Markdown notebook performs a Species Distribution Model (SDM) for *Atelopus ignescens* (the Jambato toad) using the biomod2 package. The workflow involves data preparation, pseudo-absence generation, Generalized Linear Model (GLM) fitting, and projecting the habitat suitability map.

This notebook works with relative pathways, meaning that the following structure for storing the data should be followed. For that the 'here' package is needed.

```         
root/                         <- main project folder (the root)
├─ 1985-1990/                 <- period folder
│  ├─ scripts/                <- R scripts specific to 1985–1990 (or runs for this period)
│  ├─ input/                  <- raw inputs for this period
│  ├─ output/                 <- processed outputs for this period
│  └─ figures/                <- plots/maps for this period
└─ 2018-2023/                 <- period folder
   ├─ scripts/                <- R scripts specific to 2018–2023 (or runs for this period)
   ├─ input/                  <- raw inputs for this period
   ├─ output/                 <- processed outputs for this period
   └─ figures/                <- plots/maps for this period
```

1.  **Setup and Environment**

    This chunk loads all necessary libraries and sets up the working directory

```{r}

## Load required libraries
require("terra")     # For handling spatial rasters and vectors
require("raster")    # (Legacy: Often still needed for compatibility with older spatial packages like biomod2)
require("sf")        # For simple features (vector spatial data)
require("biomod2")   # Core package for SDM
require("ggplot2")   # For enhanced plotting
require("dplyr")     # For data manipulation
require("tidyterra") # For plotting terra objects with ggplot2
require("ggtext")    # For advanced text formatting in ggplot2
require("here")      # Relative paths in this file
require("rgbif")     # GBIF R-pakage to pull occurence data from there.
require("lubridate") # package to edit data data formats.

## Set the working directory
# NOTE: Replace this path with your actual project directory.
```

# 1. Data Loading and Preparation

This section loads the species occurrence data directly from GBIF and the environmental predictor variables (rasters with the same extension, CRS and resolution).

### 1.1. Pulling GBIF occurrences

First, the taxonomic name of Atelopus must be confirmed.

```{r}
name_suggest(q="Atelopus ignescens", rank = "species")
```

Following, there is a check on the number of records available for the species, of which only those with coordinates are selected. At the end a dataframe called 'gbif_ai' is created. 'limit' is set up up to 2000 as GBIF contains 1637 records. Otherwise only 500 are retreived by default.

```{r}
gbif_ai <- occ_search(
              scientificName = "Atelopus ignescens",  # confirmed name of the species
              hasCoordinate=T,                        # logical argument to choose occ. with coordinates
              limit= 2000                             
              )

gbif_ai <- gbif_ai$data
```

### 1.2 Adding Alianza Jambato observations into the dataframe.

This dataset contains all the distribution points + collected over 2021-2022 by Alianza Jambato, Mateo Vega and Amanda Quezada.

```{r}
#project structure folder:
proj_root <- '/Alianza_Jambato_Git/Project/Atelopus niche modelling/single_SDM'

aj_ai <- read.csv(here("Project", "Atelopus niche modelling", "single_SDM",
                       "2018 - 2023", "occ_2018_2023.csv"))
aj_ai$countryCode = "EC"
aj_ai$stateProvince = "Cotopaxi"
```

```{r}
dfs = list(gbif = gbif_ai, df1 = aj_ai)

# creating a column indicating the df source
for (nm in names(dfs)) {
  dfs[[nm]]$source <- nm
}

# merging dfs by following columns in gbif_ai
all_cols = Reduce(union, lapply(dfs, names)) 

# a function to add missing colums and order them consistently
standardise_cols <- function(x, all_cols) {
  nr <- nrow(x)
  missing <- setdiff(all_cols, names(x))
  if(length(missing) > 0) {
    for (m in missing) x[[m]] <- rep(NA,nr) 
  }
  x[all_cols] # same order
}
df_ai <- lapply(dfs, standardise_cols, all_cols = all_cols)

# Combining dfs
dfaig <- do.call(rbind, df_ai)

# (optional) reset row names
row.names(dfaig) <- NULL
```

### 1.3. Checking duplicate coordinates and coordinates that are missplaced

Is always good practice to curate occurrence records from GBIF, as many of them could have innacurate identification of the corresponding species, or due to a lack of curation from the source (record hasn't been updated at the hosting institution, often a Natural History Museum). For that, we first use 'CoordinateCleaner' to clean GBIF data problems (e.g. point at province/states centroids) and finally a visual inspection based on literature review

```{r}
library("CoordinateCleaner")
library("countrycode")

# the package only works with ISO3c country codes (i.e ECU) and gfib is ISO2c (i.e. EC) thus it must be changed.

dfaig$countryCode <- countrycode(dfaig$countryCode,
                                 origin = "iso2c",
                                 destination = "iso3c")

dfaig_clean <- clean_coordinates(dfaig,
                                 lon = "decimalLongitude",
                                 lat = "decimalLatitude",
                                 countries = "countryCode",
                                 tests = c("centroids", "duplicates", "outliers"))
```

```{r}
plot(dfaig_clean)
```

```{r}
dfaig_cl <- dfaig[dfaig_clean$.summary,]
```

A few registers come from INaturalist and for protection reasons these are with a 30 km uncertainty. Thus, they need also to be filtered

```{r}
w <- "www.inaturalist.org"
m <- as.matrix(dfaig_cl)

rows <- apply(dfaig_cl, 1, function(r) any(grepl(w, r, ignore.case = TRUE)))
dfaig_cl <- dfaig_cl[!rows, ]

# there is an odd point to the east of napo tha twe are going to take out as well filtering for points -77.9

dfaig_cl <- dfaig_cl %>%
  filter(decimalLongitude <= -77.9)
```

Lastly, occurences are comparted to the known literature by exploring them on a map

```{r}
library(geodata) 
```

```{r}
# GDAM Admin-1 (provinces) for Ecuador
ec <- geodata::gadm(country = "ECU", level = 1, path = "gadm_cache")

# Excluding Galapagos
ec_mainland <- ec[!grepl("Galápagos", ec$NAME_1, ignore.case = TRUE), ]

# choosing provinces with presence of A. ignescens based on BioWeb
prov <- c("Imbabura", "Chimborazo", "Cotopaxi", "Pichincha", "Tungurahua", "Napo", "Bolivar")
ai_prov <- ec[ec$NAME_1 %in% prov, ]
plot(ai_prov)
```

```{r}
plot(ec_mainland)
occ <- vect(dfaig_cl, geom = c("decimalLongitude", "decimalLatitude"), crs = "EPSG:4326")
occ <- project(occ, crs(ec_mainland))
plot(occ, add = TRUE)
```

filtering occurences based on provinces precense

```{r}
# CRS must match 
occ <- project(occ, crs(ai_prov))

# Keep only points inside the selected provinces
occ_sel <- terra::intersect(occ, ai_prov)
```

Cheking new distribution

```{r}
plot(ec_mainland)
plot(occ_sel, add = TRUE)
```

# 2. Defining the environmental predictors

### 2.1 calling environmental and land use datasets

These datasets have been already pre-processed. They all come from Chelsa Monthly Climate Data while Land Use comes from MapBIOMAS Ecuador 3.0. DEM has been extracted from LandSat (Confirm this with Dome).

A presing problem is that tiff files are not categorical, meaning that the interpretation of the results are not possible under this way. MapBiomas, unfortunately provides maps that are characterised as continuous categories, or at least that is how Terra and QGiS interprets them. Thus, they require a prior step before entering to the loop, as the numbers must be turned into categories, and named according its [documentation](https://ecuador.mapbiomas.org/wp-content/uploads/sites/7/2025/10/ATBD-Ecuador-Coleccion-3.0-ES.pdf#page=20.75 "MapBiomas Ecuador documentation").

here is the code:

```{r}
# calling raster
lu <- rast(here("Project", "Atelopus niche modelling", "single_SDM", "2018 - 2023", "aligned_modal_18_23_crop.tif"))

# extracting pixel values from tiff to object vals
vals = sort(unique(values(lu))) 

# entering category name for each of the values within the raster following MapBiomas Documentation
labs <- c("Forest", "Open forest", "Flooded forest", "Silviculture",
          "Non-Forest Wetland", "Grassland", "Other Non-Forest Natural formation",
          "Mosaics of Uses", "Antropic non-vegetated areas", "Not observed",
          "Rocky outcrop", "Mining", "Aquaculture", "River, lake and ocean", "Glacier")

# Transforming continuous values to categories (as.factor)
lu_cat <- as.factor(lu)

# creating dataframe with names for each value
lev <- data.frame(ID = vals, land_cover = labs)

# incorporating this to GEOtiff
levels(lu_cat) <- lev

# Writing the resulting tiff file
writeRaster(lu_cat, here("Project", "Atelopus niche modelling", "single_SDM", "2018 - 2023", "aligned_LU_18_23_crop.tif"), overwrite = TRUE, wopt = list(gdal = c("COMPRESS=LZW")))

```

```{r}
### 2.1 Define Environmental Predictors

# List of paths to the environmental variables (rasters)
myExpl <- here("Project", "Atelopus niche modelling", "single_SDM", "2018 - 2023" ,c(
                               "aligned_tmax_2018_end.tif", # Max. Temperature
                               "aligned_tmin_2018_end.tif", # Min. Temperature
                               "aligned_pr_2018_end.tif", # Precipitation
                               "aligned_LU_18_23_crop.tif"#, # Land Use/Cover
                               #"aligned_DEM_end.tif" # Digital Elevation Model
                               )
)

# Load the environmental rasters into a SpatRaster object
myExpl <- rast(myExpl)
```

### 2.2 Adapting out dataset to a table readable for BIOMOD2

This table contains what are called "true presences", which simply are the pixels/grid cells where the species has been confirmed to be found.

```{r}
# converting occ_sel SpatVector class back into dataframe for BIOMOD2
occ_seldf <- as.data.frame(occ_sel, geom = "XY")
names(occ_seldf)[names(occ_seldf) == "x"] <- "decimalLongitude"
names(occ_seldf)[names(occ_seldf) == "y"] <- "decimalLatitude"

# Extract coordinates for biomod2
myRespXY <- occ_seldf[, c("decimalLongitude", "decimalLatitude")]

# Define the species presence data (all are '1' for presence points)
myResp <- rep(1, nrow(myRespXY))

# Define the species name
myRespName <- 'Atelopus_ignescens'
```

### 2.2 Formatting data to enable BIOMOD2 modelling

Often in monitoring, one aspect that is overlooked are species absences. Typically one assumes that if it is not found, it is not present, yet with no confirmed information on this aspect. This is important because in the modelling exercise, the presence/absense of one species is directly linked to the climate/environmental conditions within the grid cell used, enabling the prediction of presences in cells where similar environmental conditions are met. As we lack of true absence databases, we need to produce a pseudoabsences dataset along the territory mentioned.

#### NOTE: BIOMOD Pseudoabcenses

Often in monitoring, one aspect that is overlooked are species absences. Typically one assumes that if it is not found, it is not present, yet with no confirmed information on this aspect. This is important because in the modelling exercise, the presence/absense of one species is directly linked to the climate/environmental conditions within the grid cell used, enabling the prediction of presences in cells where similar environmental conditions are met. As we lack of true absence databases, we need to produce a pseudoabsences dataset along the territory mentioned.

The `BIOMOD_FormatingData` function is crucial. It combines the species data and environmental variables, and here, it also generates **Pseudo-Absences (PAs)**, which are necessary for presence-only models like GLM.

-   **`PA.nb.rep = 2`**: Generate 2 sets of pseudo-absences.

-   **`PA.nb.absences = 150`**: Each set will have 150 pseudo-absence points.

-   **`PA.strategy = 'random'`**: Randomly selects the locations for pseudo-absences across the study area.

    Species Distribution Models require both **Presence** (where the species is) and **Absence** (where the species is not) data. Since it's often impossible to confirm true absences, SDMs typically rely on generating **Pseudo-Absences (PAs)**.

    The parameter `PA.strategy` controls how these PA points are placed across your study area.

    |  |  |  |
    |------------------------|------------------------|------------------------|
    | **Strategy** | **PA.strategy = '...'** | **Description** |
    | **Random** | `'random'` | This is the strategy you used. PAs are selected **randomly** across the entire extent of the study area (the environmental rasters). It assumes that all unsampled areas are equally likely to be absences. |
    | **User-Defined** | `'user.defined'` | PAs are provided by the user in an external file. This is useful if you have specific knowledge about areas where the species truly does not occur (e.g., areas outside its physiological limits). |
    | **Disk** | `'disk'` | PAs are generated **outside a defined buffer/radius** around the presence points. The idea is to avoid areas very close to known presences, which are more likely to also be presences. |
    | **SRE** | `'sre'` | PAs are generated **outside the predicted suitable habitat** from a preliminary Simple Range Envelope (SRE) model. This focuses the absences in areas predicted to be unsuitable. |
    | **All** | `'all'` | Generates all possible PA datasets: one random, one user-defined (if provided), and one disk. This is usually used for ensemble modeling where you want to test the sensitivity of the final model to different PA choices. |

**Random** is a common starting point but can introduce bias if areas you didn't sample are actually suitable habitat. For **wide-ranging** or **poorly sampled** species, it's often a safe, non-restrictive option.

-   **`filter.raster = TRUE`**: Ensures occurrence points fall within the extent of the environmental rasters.

```{r}
# creating BIOMOD object

t0 <- Sys.time()

ai_biomod <- BIOMOD_FormatingData(resp.name = myRespName,
                                  resp.var = myResp,
                                  resp.xy = myRespXY,
                                  expl.var = myExpl,
                                  #Estimating pseudoabsences
                                  PA.nb.rep      = 3,            # 3 PA replicates
                                  PA.nb.absences = c(100,200, 400),         # 1000 pseudo-absences each
                                  PA.strategy    = "random",
                                  PA.dist.min    = 4000,         # e.g., 5 km
                                  PA.dist.max    = 20000,        # e.g., 50 km
                                  filter.raster  = TRUE 
                                  )

t0 <- Sys.time()

ai_biomod
```

```{r}
plot(ai_biomod)
```

# 3. Cross-validation of the datasets.

This step is needed "to assess how much of the model is stable (low variability of results when changing the data), accurate (for predictions) and able to predict new data (to flag problems like overfitting or selection bias)." (biomod2 [website](https://biomodhub.github.io/biomod2/articles/vignette_crossValidation.html) and references therein). Parameters are:\
\
**`CV.strategy = 'random'`**: Uses a random cross-validation approach.

-   **`CV.nb.rep = 1`**: Performs 1 cross-validation run (this is usually higher, e.g., 5 or 10, for more robust evaluation, but set to 1 here).

-   **`CV.perc = 0.8`**: Uses 80% of the data for training, 20% for testing.

```{r}
cv.r <- bm_CrossValidation(bm.format = ai_biomod,
                           strategy = "random",
                           nb.rep = 3,
                           k = 0.8)

print(cv.r)
summary(ai_biomod, calib.lines = cv.r)

```

# 4. Single-species Model Fitting (GLM)

### 4.1. Parametrizing the models before running.

```{r}
# Parameters for modelling the data under a General Linear Model
opt.glm.bb <- bm_ModelingOptions(
  data.type = "binary",
  models    = "GLM",
  strategy  = "bigboss",
  bm.format = ai_biomod,
  calib.lines = cv.r
)

opt.glm.bb
```

This step fits the chosen Species Distribution Model (SDM) - in this case, a Generalized Linear Model (GLM) - using the prepared data.

-   **`models = c('GLM')`**: Specifies only the Generalized Linear Model.

-   **`metric.eval = c('TSS','ROC')`**: Specifies the evaluation metrics: True Skill Statistic (TSS) and Area Under the ROC curve (AUC/ROC).

    The metrics are calculated by comparing the model's predictions (suitable vs. unsuitable) with the actual observation data (Presence vs. Absence/Pseudo-Absence). This comparison forms the basis of a **Confusion Matrix**.

    The metrics are calculated by comparing the model's predictions (suitable vs. unsuitable) with the actual observation data (Presence vs. Absence/Pseudo-Absence). This comparison forms the basis of a **Confusion Matrix**.

    $$TSS = Sensitivity + Specificity - 1$$

    **Why use both?**

    -   **ROC/AUC** is the most widely reported metric, giving a general measure of the model's discriminatory power.

    -   **TSS** is often preferred for SDMs because it is **less sensitive to the size of the pseudo-absence dataset** compared to other metrics like Kappa, giving a more balanced assessment of both correct presences and correct absences.

```{r}

# Fit the Generalized Linear Model (GLM)

myBiomodModelOut <- BIOMOD_Modeling(bm.format = ai_biomod,
                                    modeling.id = 'GLM',          # Identifier for the modeling run
                                    CV.strategy = 'random',
                                    models = c('GLM'),           # Use only GLM
                                    CV.nb.rep = 3,
                                    CV.perc = 0.8,
                                    OPT.strategy = 'bigboss',    # Optimisation strategy
                                    var.import = 3,              # Calculate variable importance
                                    metric.eval = c('TSS','ROC'),# Evaluation metrics
                                    seed.val = 1309,               # Seed for reproducibility
                                    nb.cpu = 8                   # Number of CPUs to use for parallel processing
)

# Optional: Get the evaluation scores
ai_eval <- get_evaluations(myBiomodModelOut)
```

5.  **Projecting the Model**

    After fitting the model, we use it to predict the habitat suitability across the entire study area (the environmental rasters).

```{r}

# Project the fitted single models onto the current environmental conditions
t1 <- Sys.time()

myBiomodProj <- BIOMOD_Projection(
  bm.mod = myBiomodModelOut,
  proj.name = 'Current',          # Name for the projection folder/run
  new.env = myExpl,               # Environmental data to project onto
  models.chosen = 'all',          # Use all fitted models (only GLM here)
  metric.binary = 'all',          # Compute binary maps based on all evaluation metrics
  metric.filter = 'all',          # Apply filtering based on all evaluation metrics
  build.clamping.mask = TRUE,     # Identifies areas where environment is outside training range
  nb.cpu = 2
)

Sys.time() - t1
```

```{r}
bm_PlotEvalMean(bm.out = myBiomodModelOut)
```

6.  **Save and Visualize Results**

    The final predicted habitat suitability raster is extracted and saved to a TIFF file.

```{r}

### 6.1 Extract Prediction Raster
# Get the predicted probability raster (usually a 'SpatRaster' or 'RasterStack')
# This extracts the *ensemble* of the model runs.
proj_raster <- get_predictions(myBiomodProj)

### 6.2 Save the Raster
# Save the projected habitat suitability map to a file
# NOTE: The name 'sdm_only_chelsea.tif' suggests this is a GLM run using the Chelsea climate data.
writeRaster(proj_raster, here("outputs", "ai_1985-1990_multiband.tif"),
            overwrite = TRUE)

```

```{r}

### 6.3 (Optional) Visualize the Output

# Convert to SpatRaster if necessary (check the output class of proj_raster)
# If proj_raster is a SpatRaster, you can plot it directly:
# plot(proj_raster) 

# For ggplot2 visualization (using tidyterra for easy handling)
# Check the layer names, often the output is a SpatRaster with multiple layers 
# corresponding to the PA replicates and the model. 
# You might want to plot an averaged layer if multiple are present.
if (nlyr(proj_raster) > 1) {
  # Take the mean of the replicated model predictions
  mean_proj_raster <- mean(proj_raster)
  plot_data <- mean_proj_raster
} else {
  plot_data <- proj_raster
}

# Simple visualization using tidyterra/ggplot2
ggplot() +
  geom_spatraster(data = plot_data) +
  scale_fill_viridis_c(
    option = "plasma", 
    na.value = "transparent",
    name = "Suitability (0-1000)" # biomod2 predictions are often scaled to 0-1000
  ) +
  coord_sf() +
  labs(
    title = "**Predicted Habitat Suitability for *Atelopus ignescens* (GLM)**",
    subtitle = "Based on Environmental Variables and Generated Pseudo-Absences",
    caption = "Data source: BIOMOD2 Modeling"
  ) +
  theme_minimal() +
  theme(
    plot.title = element_markdown(),
    legend.position = "bottom"
  )
```

7.  **Load and Save the Binary Map**

    Assuming you ran the BIOMOD_Modeling and BIOMOD_Projection steps:

```{r}

# 1. Get the evaluation metrics from your model object
# This table contains the "cutoff" (threshold) values for TSS and ROC
evals <- get_evaluations(myBiomodModelOut)

# 2. Extract the optimal Threshold (Cutoff) for TSS
# We filter the results to find the row where metric is "TSS" and take the cutoff value.

tss_cutoff <- evals[evals$metric.eval == "TSS", "cutoff"]

# 3. Ver el resumen estadístico de los cutoffs
summary(tss_cutoff)

# 4. Calcular la desviación estándar para ver la dispersión
sd_cutoff <- sd(tss_cutoff)
print(paste("La desviación estándar de los umbrales es:", sd_cutoff))

# 5. Visualizar la distribución de los umbrales
hist(tss_cutoff, 
     main = "Distribución de Cutoffs (TSS)", 
     xlab = "Valor de Cutoff (0-1000)", 
     col = "lightblue")

```

```{r}
prob_raster <- get_predictions(myBiomodProj)
model_names <- get_projected_models(myBiomodProj)
eval_table <- get_evaluations(myBiomodModelOut)

# 2. Extraer cutoffs asegurando coincidencia exacta
# Usamos as.numeric() para limpiar cualquier estructura extraña
current_cutoffs <- eval_table[match(model_names, eval_table$full.name) & 
                              eval_table$metric.eval == "TSS", "cutoff"]

current_cutoffs <- as.numeric(current_cutoffs)

# 3. Verificación de seguridad (Si esto falla, el match no funcionó)
if(length(current_cutoffs) != nlyr(prob_raster)) {
  stop("El número de umbrales no coincide con el número de capas del raster.")
}

# 4. Transformación Binaria
myBiomodBinaryProj <- bm_BinaryTransformation(data = prob_raster, 
                                              threshold = current_cutoffs)


# Guardar todos los mapas binarios individuales en un solo archivo multibanda
writeRaster(myBiomodBinaryProj, 
            filename = "C:/Wspace/Gabriela/SDM/1985_1990/Mapas_Binarios_Individuales_v2.tif", 
            overwrite = TRUE,
            datatype = "INT2S") # Al ser datos binarios (0 y 1), este formato ahorra espacio

```

```{r}
print(tss_cutoff)
```

8.  Mapa consensus

```{r}
# 1. Sumar todas las capas (esto da valores de 0 a 31)
suma_presencias <- sum(rast("C:/Wspace/Gabriela/SDM/1985_1990//Mapas_Binarios_Individuales_v2.tif"))

# 2. Calcular el valor de corte para el 80% (25 modelos)
umbral_consenso <- 25

# 3. Crear el mapa de consenso final
# Los pixeles que tengan 25 o más modelos a favor serán 1, el resto 0
mapa_consenso_80 <- suma_presencias >= umbral_consenso

# 4. Guardar el resultado
writeRaster(mapa_consenso_80, 
            "C:/Wspace/Gabriela/SDM/1985_1990/Consenso_80_8590_v2.tif", 
            overwrite = TRUE)
```
