---
title: "Mark-recapture analysis of the standardised monitoring of *Atelopus ignescens*"
author: "Camila Freire & Andrés Mármol-Guijarro"
date: "2026-01-20"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)

library(tidyverse)
library(lubridate)
library(janitor)
library(stringr)
library(readr)
library(scales)
library(here)
library(RMark)
library(knitr)
library(kableExtra)
library(gt)
library(officer)
library(flextable)

# Shared paths
input_dir  <- here("Project", "Mark recapture", "input_data")
output_dir <- here("Project", "Mark recapture", "output")
```

## Data loading

```{r load-data}
df <- read.csv(file.path(input_dir, "ai_dataset.csv"), header = TRUE)

head(df, 2)
```

## Data cleaning

```{r clean-names}
df_clean <- janitor::clean_names(df)

names(df_clean)
```

### Standardise sex codes and parse dates

```{r clean-sex-dates}
# Standardise sex codes to M / F / I / U
df_clean <- df_clean |>
  mutate(
    sexo = toupper(sexo),
    sexo = dplyr::recode(sexo,
                         "M" = "M",   # male
                         "H" = "F",   # female
                         "I" = "I",   # undetermined
                         .default = "U")
  )

# Parse fecha from character to Date, trying multiple formats
df_clean <- df_clean |>
  mutate(
    fecha_parsed = parse_date_time(
      fecha,
      orders = c("Y-m-d", "Y/m/d", "d/m/Y", "d-m-Y", "m/d/Y", "m-d-Y"),
      tz = "UTC"
    ) |> as_date()
  )

summary(df_clean$fecha_parsed)
```

## Building the capture history

### Collapse to individual × month

```{r individual-month}
df_month <- df_clean |>
  filter(!is.na(id), !is.na(fecha_parsed)) |>
  mutate(month = floor_date(fecha_parsed, "month")) |>
  distinct(id, month)

head(df_month)
```

### Full sequence of months in the study

```{r month-sequence}
# Exclude December 2024 (no sampling that month)
all_months <- tibble(
  month = seq(min(df_month$month), max(df_month$month), by = "month")
) |>
  filter(!(year(month) == 2024 & month(month) == 12))
```

### Presence / absence matrix

```{r presence-absence}
# Ensure every id has a row for every month; mark presence (1) / absence (0)
det_full <- expand_grid(id = unique(df_month$id), month = all_months$month) |>
  left_join(df_month |> mutate(present = 1L), by = c("id", "month")) |>
  mutate(present = replace_na(present, 0L))

# Pivot to wide 0/1 table
mat_df <- det_full |>
  arrange(id, month) |>
  mutate(month_lab = format(month, "%Y_%m")) |>
  select(id, month_lab, present) |>
  pivot_wider(names_from = month_lab, values_from = present, values_fill = 0L) |>
  arrange(id)

cap_matrix <- mat_df |>
  tibble::column_to_rownames("id") |>
  as.matrix()
```

### Create capture-history strings

```{r capture-history}
ch_tbl <- mat_df |>
  mutate(ch = apply(select(mat_df, -id), 1, paste0, collapse = "")) |>
  select(id, ch)
```

### Extract individual covariates

```{r covariates}
# One sex and age record per id (first non-missing observation)
covs <- df_clean |>
  select(id, sexo, edad) |>
  group_by(id) |>
  summarise(
    sexo = dplyr::first(sexo[!is.na(sexo)], default = NA_character_),
    edad = dplyr::first(edad[!is.na(edad)], default = NA_character_),
    .groups = "drop"
  )

# Final RMark input table
rmark_df <- ch_tbl |>
  left_join(covs, by = "id") |>
  mutate(freq = 1L)
```

## POPAN open-population model

### Process data and build design matrix

```{r process-data}
aipop.process <- process.data(
  rmark_df,
  model         = "POPAN",
  time.interval = c(rep(1, 15), 2, rep(1, 5))
)

aipop.ddl <- make.design.data(aipop.process)

aipop.ddl
```

### Define and run candidate models

Three candidate models are compared, crossing constant vs. time-varying
formulations for survival (Phi), detection probability (p), and entry
probability (pent). Superpopulation size (N) is always constant.

```{r define-models}
run.ai <- function() {

  # Survival (Phi)
  Phi.dot  <- list(formula = ~1)
  Phi.time <- list(formula = ~time)

  # Detection probability (p)
  p.dot  <- list(formula = ~1)
  p.time <- list(formula = ~time)

  # Entry probability (pent)
  pent.dot  <- list(formula = ~1)
  pent.time <- list(formula = ~time)

  # Superpopulation size (N)
  N.dot <- list(formula = ~1)

  # Model 1: all parameters constant
  phi.dot.p.dot.pent.dot.N <- mark(
    aipop.process, aipop.ddl,
    model.parameters = list(Phi = Phi.dot, p = p.dot, pent = pent.dot, N = N.dot)
  )

  # Model 2: Phi and pent time-varying, p constant
  phi.time.p.dot.pent.time.N <- mark(
    aipop.process, aipop.ddl,
    model.parameters = list(Phi = Phi.time, p = p.dot, pent = pent.time, N = N.dot)
  )

  # Model 3: all parameters time-varying
  phi.time.p.time.pent.time.N <- mark(
    aipop.process, aipop.ddl,
    model.parameters = list(Phi = Phi.time, p = p.time, pent = pent.time, N = N.dot)
  )

  return(collect.models())
}
```

```{r run-models, results='hide', message=FALSE}
ai.results <- run.ai()
```

### Model selection table

```{r model-table}
options(width = 150)
ai.results
```

## Results from the top model

The top model is `phi.time.p.dot.pent.time.N`, in which survival and entry
probability vary over time but detection probability is held constant. This
assumes that all sampling events were conducted by the same observers, that
weather conditions do not influence recapture probability, and that site
conditions have remained stable over the study period.

```{r top-model}
top <- ai.results$phi.time.p.dot.pent.time.N

summary(top, showall = FALSE)

round(top$results$real[, 1:4], 2)
```

### Survival probability (Phi)

```{r phi}
top.phi <- top$results$real[1:21, ]
top.phi
```

### Detection probability (p)

```{r p}
top.p <- top$results$real[22, ]
top.p
```

### Entry probability (pent)

```{r pent}
top.pent <- top$results$real[24:44, ]
round(top.pent[, 1:4], 5)
```

### Superpopulation size (N)

```{r N}
top.N <- top$results$real[23, ]
top.N
```

## Derived abundance estimates per sampling occasion

```{r entry-proportions}
# b:  probability of entry per occasion
# bo: proportion of the population present at t1
b  <- top.pent$estimate
bo <- 1 - sum(top.pent$estimate)

round(c(bo, b), 4)
```

```{r abundance-per-occasion}
# Number of new entrants per occasion: B.occ = b × N
B.occ     <- b * top.N$estimate
B.occ_inf <- b * top.N$estimate - b * top.N$se
B.occ_upp <- b * top.N$estimate + b * top.N$se

# Initial cohort at t1: B0 = bo × N
B0     <- bo * top.N$estimate
B0_inf <- bo * top.N$estimate - bo * top.N$se
B0_upp <- bo * top.N$estimate + bo * top.N$se

# N per sampling occasion
N_occ <- c(B0, B.occ)
N_inf <- c(B0_inf, B.occ_inf)
N_upp <- c(B0_upp, B.occ_upp)
```

```{r plot-dataframe}
# Assemble all derived parameters into a single plot dataframe
survival <- c(NA, top.phi$estimate)
s_inf    <- c(NA, top.phi$estimate - top.phi$se)
s_upp    <- c(NA, top.phi$estimate + top.phi$se)

entry <- c(NA, top.pent$estimate)
e_inf <- c(NA, top.pent$estimate - top.pent$se)
e_upp <- c(NA, top.pent$estimate + top.pent$se)

df_plot <- data.frame(N_occ, N_inf, N_upp, survival, s_inf, s_upp, entry, e_inf, e_upp)

# Insert NA row to represent the 2-month gap between occasions 16 and 17
na_row        <- as.data.frame(as.list(rep(NA, ncol(df_plot))))
names(na_row) <- names(df_plot)

df_plot2           <- rbind(df_plot[1:16, ], na_row, df_plot[17:nrow(df_plot), ])
rownames(df_plot2) <- NULL

round(df_plot2, 1)

# Assign monthly dates starting August 2023
df_plot2$month_year <- seq(
  from       = as.Date("2023-08-01"),
  by         = "month",
  length.out = nrow(df_plot2)
)
```

## Figures

### Individuals per sampling occasion vs. survival probability

```{r plot-survival}
scale_factor <- quantile(df_plot2$N_occ, 0.95, na.rm = TRUE) /
                quantile(df_plot2$survival, 0.95, na.rm = TRUE)

ylim_left  <- c(0, max(df_plot2$N_upp, na.rm = TRUE) * 1.1)
ylim_right <- c(0, max(df_plot2$survival, na.rm = TRUE) * 1.1)

ggplot(df_plot2, aes(x = month_year)) +

  # Blue bars: estimated individuals
  geom_col(aes(y = N_occ),
           fill = "#0072B2", color = "#0072B2",
           linewidth = 0.1, width = 6.5, alpha = 0.55, na.rm = TRUE) +
  geom_errorbar(aes(ymin = N_inf, ymax = N_upp),
                width = 0.15, color = "#002244", linewidth = 0.5, na.rm = TRUE) +

  # Orange line: survival probability
  geom_errorbar(aes(ymin = s_inf * scale_factor, ymax = s_upp * scale_factor),
                width = 0.2, color = "#D55E00", na.rm = TRUE) +
  geom_point(aes(y = survival * scale_factor),
             color = "#D55E00", size = 3, na.rm = TRUE) +
  geom_line(aes(y = survival * scale_factor),
            color = "#D55E00", linewidth = 1, na.rm = TRUE) +

  geom_vline(xintercept = as.Date("2024-09-01"),
             color = "#D55E00", linetype = "dashed", linewidth = 1) +

  scale_x_date(date_breaks = "1 month", date_labels = "%b-%Y",
               expand = c(0.01, 0.01)) +
  scale_y_continuous(
    name     = "Individuals per sampling event",
    limits   = ylim_left,
    sec.axis = sec_axis(~ . / scale_factor,
                        name   = "Prob. survival to next sampling event",
                        breaks = seq(0, max(ylim_right), length.out = 5))
  ) +
  theme_minimal(base_size = 13) +
  theme(
    axis.title.y       = element_text(color = "#0072B2", size = 13, face = "bold"),
    axis.title.y.right = element_text(color = "#D55E00", size = 13, face = "bold"),
    axis.text.x        = element_text(angle = 90, vjust = 0.5, hjust = 1, size = 10),
    axis.text.y        = element_text(size = 10, color = "#0072B2"),
    axis.text.y.right  = element_text(size = 10, color = "#D55E00"),
    panel.grid.major   = element_line(color = "gray85", linewidth = 0.5),
    panel.grid.minor   = element_line(color = "gray90", linewidth = 0.3),
    axis.line          = element_line(color = "black", linewidth = 0.6),
    panel.background   = element_rect(fill = "gray97", color = NA),
    plot.background    = element_rect(fill = "white", color = NA)
  )
```

### Load meteorological data

```{r load-met}
df_prec <- read_csv(file.path(input_dir, "Prec.csv")) |>
  mutate(mes = as.Date(mes))

df_temp <- read_csv(file.path(input_dir, "Temp.csv")) |>
  mutate(mes = as.Date(mes))

max_occ  <- max(df_plot2$N_occ, na.rm = TRUE)
max_prec <- max(df_prec$precipitacion_total, na.rm = TRUE)
max_temp <- max(df_temp$temp_med, na.rm = TRUE)
```

### Individuals vs. precipitation (exploratory)

```{r plot-met-exploratory}
scale_prec <- max_occ / max_prec
scale_temp <- max_occ / max_temp

ggplot() +

  # Blue line: individuals
  geom_errorbar(data = df_plot2,
                aes(x = month_year, ymin = N_inf, ymax = N_upp),
                width = 0.2, color = "#0072B2", na.rm = TRUE) +
  geom_point(data = df_plot2, aes(x = month_year, y = N_occ),
             color = "#0072B2", size = 3, na.rm = TRUE) +
  geom_line(data = df_plot2, aes(x = month_year, y = N_occ),
            color = "#0072B2", linewidth = 1, na.rm = TRUE) +

  # Green line: precipitation
  geom_line(data = df_prec,
            aes(x = mes, y = precipitacion_total * scale_prec),
            color = "#009E73", linewidth = 1.2, na.rm = TRUE) +
  geom_point(data = df_prec,
             aes(x = mes, y = precipitacion_total * scale_prec),
             color = "#009E73", size = 3, na.rm = TRUE) +

  # Orange line: temperature
  geom_line(data = df_temp,
            aes(x = mes, y = temp_med * scale_temp),
            color = "#D55E00", linewidth = 1.2, na.rm = TRUE) +
  geom_point(data = df_temp,
             aes(x = mes, y = temp_med * scale_temp),
             color = "#D55E00", size = 3, na.rm = TRUE) +

  scale_x_date(date_breaks = "1 month", date_labels = "%b-%Y",
               expand = c(0.01, 0.01)) +
  scale_y_continuous(
    name     = "Individuos por evento de muestreo",
    sec.axis = sec_axis(~ . / scale_prec,
                        name = "Precipitación (mm) / Temperatura (°C)")
  ) +
  theme_minimal(base_size = 13) +
  theme(
    axis.title.y       = element_text(color = "#0072B2", size = 13, face = "bold"),
    axis.title.y.right = element_text(color = "black", size = 13, face = "bold"),
    axis.text.x        = element_text(angle = 90, vjust = 0.5, hjust = 1, size = 10),
    axis.text.y        = element_text(size = 10, color = "#0072B2"),
    axis.text.y.right  = element_text(size = 10),
    panel.grid.major   = element_line(color = "gray85", linewidth = 0.5),
    panel.grid.minor   = element_line(color = "gray90", linewidth = 0.3),
    axis.line          = element_line(color = "black", linewidth = 0.6),
    panel.background   = element_rect(fill = "gray97", color = NA),
    plot.background    = element_rect(fill = "white", color = NA)
  )
```

### Individuals vs. temperature

```{r plot-temperature}
scale_temp <- max_occ / max(df_temp$temp_med1, na.rm = TRUE)

plottem <- ggplot() +

  # Blue bars: estimated individuals
  geom_bar(data = df_plot2,
           aes(x = month_year, y = N_occ),
           stat = "identity", fill = "#0072B2", color = "#0072B2", width = 15) +
  geom_errorbar(data = df_plot2,
                aes(x = month_year, ymin = N_inf, ymax = N_upp),
                width = 0.2, color = "black") +

  # Orange line: temperature
  geom_line(data = df_temp,
            aes(x = mes, y = temp_med * scale_temp),
            color = "#D55E00", linewidth = 1.2) +
  geom_point(data = df_temp,
             aes(x = mes, y = temp_med * scale_temp),
             color = "#D55E00", size = 3) +

  scale_x_date(date_breaks = "1 month", date_labels = "%b-%Y",
               expand = c(0.01, 0.01)) +
  scale_y_continuous(
    name     = "Individuos por evento de muestreo",
    breaks   = seq(0, 30, by = 5),
    sec.axis = sec_axis(~ . / scale_temp, name = "Temperatura promedio (°C)")
  ) +
  theme_minimal(base_size = 13) +
  theme(
    axis.title.y       = element_text(color = "#0072B2", face = "bold"),
    axis.title.y.right = element_text(color = "#D55E00", face = "bold"),
    axis.text.x        = element_text(angle = 90, vjust = 0.5, hjust = 1),
    panel.grid.major   = element_line(color = "gray85"),
    panel.grid.minor   = element_line(color = "gray90"),
    panel.background   = element_rect(fill = "gray97", color = NA),
    axis.line.x        = element_line(color = "black", linewidth = 1),
    axis.line.y        = element_line(color = "black", linewidth = 1),
    axis.ticks.x       = element_line(color = "black", linewidth = 0.6),
    axis.ticks.y       = element_line(color = "black", linewidth = 0.6)
  )

ggsave(
  filename = file.path(output_dir, "Grafica_individuos_tem.png"),
  plot = plottem, width = 10, height = 6, dpi = 300
)

plottem
```

### Combined: individuals, precipitation and temperature

```{r plot-met-combined}
scale_temp <- max_prec / max_temp
scale_occ  <- max_prec / max_occ

plot_met <- ggplot() +

  # Precipitation bars
  geom_col(data = df_prec,
           aes(x = mes, y = precipitacion_total, fill = "Precipitación (mm)"),
           alpha = 0.6) +

  # Temperature line
  geom_line(data = df_temp,
            aes(x = mes, y = temp_med * scale_temp, color = "Temperatura (°C)"),
            linewidth = 1.3) +
  geom_point(data = df_temp,
             aes(x = mes, y = temp_med * scale_temp, color = "Temperatura (°C)"),
             size = 3) +

  # Individuals line
  geom_line(data = df_plot2,
            aes(x = month_year, y = N_occ * scale_occ, color = "Individuos"),
            linewidth = 1.3) +
  geom_point(data = df_plot2,
             aes(x = month_year, y = N_occ * scale_occ, color = "Individuos"),
             size = 3) +

  geom_vline(xintercept = as.Date("2024-09-01"),
             color = "red", linetype = "dashed", linewidth = 1) +

  scale_x_date(date_breaks = "1 month", date_labels = "%b-%Y",
               expand = c(0.01, 0.01)) +
  scale_y_continuous(
    name     = "Precipitación total (mm)",
    sec.axis = sec_axis(~ . / scale_temp,
                        name = "Temperatura (°C) / Individuos por muestreo")
  ) +
  scale_fill_manual(values  = c("Precipitación (mm)" = "#56B4E9"), name = NULL) +
  scale_color_manual(values = c("Temperatura (°C)" = "#D55E00",
                                "Individuos"       = "#0072B2"), name = NULL) +
  theme_minimal(base_size = 13) +
  theme(
    axis.title.y       = element_text(color = "black", face = "bold"),
    axis.title.y.right = element_text(color = "black", face = "bold"),
    axis.text.x        = element_text(angle = 90, vjust = 0.5, hjust = 1),
    axis.text.y        = element_text(color = "black"),
    axis.text.y.right  = element_text(color = "black"),
    panel.grid.major   = element_line(color = "gray85"),
    panel.grid.minor   = element_line(color = "gray90"),
    panel.background   = element_rect(fill = "gray97", color = NA),
    plot.background    = element_rect(fill = "white", color = NA),
    axis.line          = element_line(color = "black", linewidth = 0.6),
    legend.position    = "right",
    legend.box         = "vertical",
    legend.text        = element_text(size = 12)
  )

ggsave(
  filename = file.path(output_dir, "Grafica_individuos_met_f.png"),
  plot = plot_met, width = 10, height = 6, dpi = 300
)

plot_met
```

## Descriptive capture summary

### Prepare capture table

```{r capture-summary}
# Re-process df for cross-month recapture tagging
df <- df |>
  clean_names() |>
  mutate(
    fecha      = parse_date_time(fecha, orders = c("dmy", "ymd", "mdy")),
    month_date = floor_date(as_date(fecha), "month"),
    sexo       = str_squish(str_to_lower(sexo)),
    sex_grp    = case_when(
      str_starts(sexo, "m") ~ "male",
      str_starts(sexo, "h") ~ "female",
      TRUE                  ~ "indef"
    )
  ) |>
  filter(!is.na(month_date), !is.na(id))

# Full month sequence
all_months <- tibble(
  month_date = seq(min(df$month_date), max(df$month_date), by = "month")
)

# First month each individual was seen
first_seen <- df |>
  group_by(id) |>
  summarise(first_month = min(month_date), .groups = "drop")

# Tag recaptures
df_tag <- df |>
  left_join(first_seen, by = "id") |>
  mutate(is_recapture = month_date > first_month)

# Overall per-month totals
overall <- df_tag |>
  group_by(month_date) |>
  summarise(
    total_captures   = n(),
    total_recaptures = n_distinct(id[is_recapture]),
    .groups = "drop"
  )

# Per-sex captures & recaptures, wide layout
by_sex <- df_tag |>
  group_by(month_date, sex_grp) |>
  summarise(
    captures   = n(),
    recaptures = n_distinct(id[is_recapture]),
    .groups = "drop"
  ) |>
  pivot_wider(
    names_from  = sex_grp,
    values_from = c(captures, recaptures),
    values_fill = 0
  )

# Combine, fill zero months, add month label
summary_tbl <- all_months |>
  left_join(overall, by = "month_date") |>
  left_join(by_sex,  by = "month_date") |>
  replace_na(list(
    total_captures    = 0, total_recaptures  = 0,
    captures_male     = 0, recaptures_male   = 0,
    captures_female   = 0, recaptures_female = 0,
    captures_indef    = 0, recaptures_indef  = 0
  )) |>
  arrange(month_date) |>
  mutate(month = format(month_date, "%b-%Y")) |>
  select(
    month,
    total_captures,   total_recaptures,
    captures_male,    recaptures_male,
    captures_female,  recaptures_female,
    captures_indef,   recaptures_indef
  )

summary_tbl
```

### Table: HTML (gt)

```{r table-gt}
summary_tbl |>
  gt() |>
  tab_header(
    title    = "Monthly captures and recaptures",
    subtitle = "By sex and category"
  ) |>
  cols_label(
    month             = "Month",
    total_captures    = "Total captures",
    total_recaptures  = "Total recaptures",
    captures_male     = "Males captured",
    recaptures_male   = "Males recaptured",
    captures_female   = "Females captured",
    recaptures_female = "Females recaptured",
    captures_indef    = "Indef. captured",
    recaptures_indef  = "Indef. recaptured"
  ) |>
  fmt_number(columns = -month, decimals = 0) |>
  tab_options(table.font.size = px(11))
```

### Table: LaTeX (kableExtra)

```{r table-latex}
summary_tbl |>
  kable(format = "latex", booktabs = TRUE,
        caption = "Monthly captures and recaptures of Atelopus sp.") |>
  kable_styling(latex_options = c("striped", "hold_position"))
```

### Export to Word

```{r table-docx}
ft <- flextable(summary_tbl)
save_as_docx(ft, path = file.path(output_dir, "summary_table_DT.docx"))
```

## Individuals (model estimate + observed captures) vs. precipitation

Add observed captures to the plot dataframe.

```{r add-observed-captures}
df_plot2$total_captures <- summary_tbl$total_captures
```

```{r plot-precipitation-final}
df_plot2 <- df_plot2 |> mutate(month_year = as.Date(month_year))
df_prec  <- df_prec  |> mutate(mes = as.Date(mes))

scale_prec <- max_occ / max(df_prec$precipitacion_total, na.rm = TRUE)

# Long format for side-by-side bars
df_bars <- df_plot2 |>
  select(month_year, N_occ, total_captures, N_inf, N_upp) |>
  pivot_longer(
    cols      = c(N_occ, total_captures),
    names_to  = "serie",
    values_to = "valor"
  ) |>
  mutate(serie = recode(serie,
                        N_occ          = "Estimación (modelo)",
                        total_captures = "Capturas (observado)"))

plotpre <- ggplot() +
  geom_col(data = df_bars,
           aes(x = month_year, y = valor, fill = serie),
           position = position_dodge(width = 20), width = 18) +
  geom_errorbar(data = df_plot2,
                aes(x = month_year, ymin = N_inf, ymax = N_upp),
                position = position_dodge(width = 20),
                width = 3, color = "black") +
  geom_line(data = df_prec,
            aes(x = mes, y = precipitacion_total * scale_prec),
            color = "#009E73", linewidth = 1.2) +
  geom_point(data = df_prec,
             aes(x = mes, y = precipitacion_total * scale_prec),
             color = "#009E73", size = 3) +
  geom_vline(xintercept = as.numeric(as.Date("2024-07-01")),
             color = "red", linetype = "dashed", linewidth = 1) +
  scale_fill_manual(values = c("Estimación (modelo)"  = "#0072B2",
                               "Capturas (observado)" = "#CF5C36")) +
  scale_x_date(date_breaks = "1 month", date_labels = "%b-%Y",
               expand = c(0.01, 0.01)) +
  scale_y_continuous(
    name     = "Individuos por evento de muestreo",
    breaks   = seq(0, 30, by = 5),
    sec.axis = sec_axis(~ . / scale_prec, name = "Precipitación total (mm)")
  ) +
  labs(x = "Mes") +
  theme_minimal(base_size = 13) +
  theme(
    axis.title.y       = element_text(color = "#0072B2", face = "bold"),
    axis.title.y.right = element_text(color = "#009E73", face = "bold"),
    axis.text.x        = element_text(angle = 90, vjust = 0.5, hjust = 1),
    panel.grid.major   = element_line(color = "gray85"),
    panel.grid.minor   = element_line(color = "gray90"),
    panel.background   = element_rect(fill = "gray97", color = NA),
    axis.line.x        = element_line(color = "black", linewidth = 1),
    axis.line.y        = element_line(color = "black", linewidth = 1),
    axis.ticks.x       = element_line(color = "black", linewidth = 0.6),
    axis.ticks.y       = element_line(color = "black", linewidth = 0.6),
    legend.position      = "top",
    legend.title         = element_blank(),
    legend.justification = "center"
  )

ggsave(
  filename = file.path(output_dir, "Grafica_individuos_pre_v2.png"),
  plot = plotpre, width = 10, height = 6, dpi = 300
)

plotpre
```
